Income's Impact on Food Spending 

### 2024-04-18 paper/paper.qmd
updated the pdf file

Over the past thirty years in the US, wages have consistently risen, while tax rates have remained relatively stable. This has resulted in people having more disposable income, which is the money left after taxes are deducted from one's income. Disposable income is crucial to consider as it reflects the actual amount of money available for spending. Studies indicate that when individuals have higher disposable income, they generally spend more [@friend1946relationship]. We are particularly interested in examining this relationship in the context of food.

@@

For a detailed breakdown of each dataset, refer to @tbl-1. Across all datasets, a few consistent features are notable. Firstly, we confined our data selection to the quarterly period spanning from 2007 to 2022. We anchored our starting point in 2007 due to the unavailability of durable/nondurable goods data prior to 2007. We were interested in a comprehensive 15-year modeling period, which culminated in 2022.

@@

FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S. Census Bureau, Bureau of Labor Statistics, and U.S. Bureau of Economic Analysis (BEA), among others. These datasets are aggregated at the country level. We use national-level data because it provides a broader perspective on general economic trends in the U.S. All five datasets we have used originate from the BEA. The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies. Upon receiving these datasets, BEA makes adjustments to the raw survey data. For instance, they fill in missing gaps in the datasets. For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from non-merchant sellers. Therefore, BEA incorporates a separate survey to fill this gap, giving us a more comprehensive view of the overall economic conditions.

FRED collaborates with both public and private organizations, such as the BEA, to populate its comprehensive database. The goal of FRED is to serve as a centralized data bank where you can find information on all facets of the U.S. economy, eliminating the need to search various agencies individually for data. However, FRED does make further adjustments to BEA datasets before releasing them. For example, when aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].

@@

The income and price elasticities of food are crucial factors in understanding why there is not much change in food spending despite changes in income. Income elasticity measures how much food demand shifts when income changes, while price elasticity assesses the impact of price changes on food demand.

>>>>

Over the past thirty years in the US, wages have consistently risen, while tax rates have remained relatively stable. This has resulted in people having more disposable income, which is the money left after taxes are deducted from one's income. Disposable income is an important factor to consider as it reflects the actual amount of money available for spending. Studies indicate that when individuals have higher disposable income, they generally spend more [@friend1946relationship]. We are particularly interested in examining this relationship in the context of food.

@@

For a detailed breakdown of each dataset, refer to @tbl-1. Across all datasets, a few consistent features are notable. Firstly, we confined our data selection to the quarterly period spanning from 2007 to 2022. We anchored our starting point in 2007 due to the unavailability of durable/nondurable goods data prior to 2007. We are interested in a 15-year modeling period, which culminated in 2022.

@@

FRED datasets, essentially time series variants, are compiled by the US Federal Reserve and sourced from government agencies like the U.S. Census Bureau, Bureau of Labor Statistics, and U.S. Bureau of Economic Analysis (BEA), among others. These datasets are aggregated at the country level. We use national-level data because it provides a broader perspective on general economic trends in the U.S. All five datasets we have used originate from the BEA. The BEA gathers data from over 360 surveys and collections sponsored by various Federal and private agencies. Upon receiving these datasets, BEA makes adjustments to the raw survey data. For instance, they fill in missing gaps in the datasets. For example, when estimating changes in private inventories for Gross Domestic Product (GDP), the primary survey is the Census Bureau’s Monthly Wholesale Trade Survey does not include inventories from non-merchant sellers. Therefore, BEA incorporates a separate survey to fill this gap, giving us a wider view of the overall economic conditions.

FRED collaborates with both public and private organizations, such as the BEA, to populate their database. The goal of FRED is to serve as a centralized data bank where you can find information on all facets of the U.S. economy, eliminating the need to search various agencies individually for data. However, FRED does make further adjustments to BEA datasets before releasing them. For example, when aggregating data through averaging, summing, or end-of-period calculations, FRED typically overlooks missing values, except for the first and last observations of each month [@fred].

@@

The income and price elasticities of food are important factors in understanding why there is not much change in food spending despite changes in income. Income elasticity measures how much food demand shifts when income changes, while price elasticity assesses the impact of price changes on food demand.
